---
title: "Report_Draft"
output: html_document
date: "2025-04-23"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This research project aims to understand whether individual's perception of economic inequality has an impact on subjective well-being. A consistent body of research (see Lynch et al., 2004; Wilkinson and Pickett, 2009) has indeed highlighted that economic inequality has an adverse impact on individuals' happiness, general health and social cohesion. Income inequality is generally measured with indicators such as the Gini index, however a recent research stream (see Vezzoli et al., 2022) has drawn attention to the much more relevant impact of perception of inequality on well-being, compared to other objective measures of inequality. 

This project test whether there is a correlation between perception of inequality and subjective well-being among people living in European countries. We hypothesize that the lower the perception of economic inequality, the higher the level of subjective well-being. We assess this hypothesis using the data-set provided by the European Social Survey (ESS), a cross-national survey across Europe which provides rigorous data on attitudes, beliefs and behavior patterns among individuals living in European countries.



## Data and methods


To measure perception of inequality and subjective well-being we use the ESS Round 9, which has a section with a special focus on justice and fairness. 

From the data-set we selected two variables that provide the most accurate measurements to operationalize our concepts of interest. 

- Variable 1: 'happy' - How happy are you

- Variable 2: 'wltdffr' - Differences in wealth in country, how fair

After the selection, we have created two new versions of the variables. They contain fewer meaningful categories in order to reduce complexity and improve the statistical power of our analysis, and no missing values.

- New variable 1: 'happy_group' - categorical variable with 3 values (Low, Medium, High)

- New variable 2: 'wealth' - categorical variable with 3 values (Fair, Somewhat fair, Very unfair)



### Anlysis


We first conducted a descriptive analysis to explore the distribution of happiness and perceived wealth fairness across the sample. We then conducted a chi-square test to assess whether the two categorical variables were statistically associated. To further investigate if the relations was robust we estimated a linear regression model with other controlling variables.


To measure the percentages of the two variables together we created a cross-table with percentages.
We found out that the majority of individuals fall into the “High” happiness category, regardless of their perception of fairness. However, the proportion of "Low" happiness is highest among those who perceive wealth as "Very Unfair" (3.0%), compared to only 0.2% in the “Fair” group.

We then conducted a chi-square test, confirming that the association between wealth and happy_group is statistically significant (χ² = 879.96, df = 4, p < 0.001), suggesting that perceptions of inequality are meaningfully related to levels of subjective wellbeing.


We also calculated the average happiness level within each category of perceived wealth fairness. The means are as follows:

 - Fair: 8.16
 - Somewhat Fair: 7.74
 - Very Unfair: 7.47

These differences, although modest, show a clear decreasing trend in average happiness as the perception of unfairness increases. This trend is accompanied by 95% confidence intervals that do not overlap substantially, indicating statistically robust group differences.



# Graphical representation


```{r}
p <- ggplot(ESS9_clean, aes(x = wealth, fill = happy_group)) +
  geom_bar(position = "fill") +
  labs(title = "Association between Wealth Discrepancies and Happiness",
       x = "Wealth Discrepancy",
       y = "Proportion of Happiness Level") +
  scale_fill_brewer(palette = "Set4") +
  theme_minimal()

print(p)
```



# ADD-ON analysis:

To assess the relationship between perceived inequality and happiness while controlling for individual-level characteristics, we estimated a linear regression model including gender (gndr), age (agea), and years of education (eduyrs) alongside perceived inequality (wltdffr). 

Results:

wltdffr; +0.052 – Positive and statistically significant

eduyrs; +0.024 – Positive and statistically significant

gndr; +0.003 – Small and not significant

agea; -0.00006 – Negligible

These results suggest that both higher perceived inequality and higher education levels are independently associated with higher reported happiness, although the overall explanatory power of the model remains low.


```{r}
p_2 <- tidy(control_var, conf.int = TRUE) |> 
  filter(term != "(Intercept)") |> 
  ggplot(aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(
    title = "Effect of predictors on happiness",
    x = "Predictors",
    y = "Estimated effect (with 95% CI)"
  ) +
  theme_minimal()

print(p_2)
```



# Results 





## Discussion 








### Bibbliography

Lynch, J., Smith, G.D., Harper, S., Hillemeier, M., Ross, N., Kaplan, G.A., Wolfson, M., 2004. Is income inequality a determinant of population health? Part 1. A systematic review. Milbank Q., 82(1), 5-99. 

Vezzoli, M., Valtorta, R.R., Mari, S., Durante, F. and Volpato, C., 2023. Effects of objective and subjective indicators of economic inequality on subjective well‐being: Underlying mechanisms. Journal of Applied Social Psychology, 53(2), 85-100.

Wilkinson, R.G., & Pickett, K.E., 2009. Income Inequality and Social Dysfunction. Annual Review of Sociology, 35, 493–511.


